---
title: "Modeling the dynamic heat inactivation of *Vibrio parahaemolyticus* in oyster digestive diverticula"
author: "Weifan Wu, et al."
date: last-modified
format:
  docx:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 7
    highlight-style: github
    code-wrap: true
editor: visual
---

# Global Environment & Session Info

```{r global_environ}
# Set Seed
set.seed(5025)
# Limit the number of significant digits to 3
options(digits = 3)

# Packages
## Data Entry and Manipulation
library(tidyverse)
library(here)
library(readxl)
library(skimr)
library(nlme)
library(fitdistrplus)
## Bioinactivation
library(bioinactivation)
library(mvtnorm) # For calc CI and PI for graphs
library(FME) # for performing global fit on dynamic data
library(cowplot)
library(emmeans)
## Data Visualization
library(patchwork)
## Session Info
library(sessioninfo)
sessioninfo::session_info (pkgs = c("attached"))
knitr::opts_chunk$set(error = TRUE, warning = FALSE, message = FALSE)

quiet_predict_inactivation <- function(...) {
  result <- NULL
  invisible(capture.output(
    result <- suppressWarnings(bioinactivation::predict_inactivation(...)),
    type = "output"
  ))
  result
}
```

# Data Import

```{r eval=FALSE, echo=FALSE}
## 1. Temperature Profiles 
df_comeup <- read_excel(here ("Data", "vibrio_oys_data_new.xlsx"), sheet = 4) %>%
  mutate(temp_target_c = case_when(temp_target_c == 100 ~ 98,
                                   TRUE ~ temp_target_c),
         temp_target_c = as.factor(temp_target_c),
         strain = as.factor(strain)
         )

saveRDS(df_comeup, here("Data", "df_comeup.RDS"))

## 2. Dynamic inactivation data
df_dyn <- read_excel(here ("Data", "vibrio_oys_data_new.xlsx"), sheet = 3) %>%
  mutate(temp_target_c = case_when(temp_target_c == 100 ~ 98,
                                   TRUE ~ temp_target_c),
         temp_target_c = as.factor(temp_target_c),
         strain = case_when(strain == "VP2" ~ "TX_2103",
                              strain == "VP3" ~ "FDA_R2"),
         strain = as.factor(strain))

saveRDS(df_dyn, here("Data", "df_dyn.RDS"))
```

```{r}
df_dyn <- readRDS (here("Data", "df_dyn.RDS"))

df_comeup <- readRDS (here("Data", "df_comeup.RDS"))

skim(df_dyn)
skim(df_comeup)
```

# Statistical comparison between two different media

```{r}
df_media <- read_excel(here("Data", "Vibrio Parahaemolyticus on PBS at 74C April 2026.xlsx"), sheet = 3) %>%
  mutate(plate = as.factor(plate)) %>%
  filter(plate != "TCBS")

# Model with strain-by-temperature fixed effect

fit_media_lm <- lm(log_cfu ~ plate * time, data = df_media)

emm <- emmeans(fit_media_lm, ~ plate)

contrast_df <- as.data.frame(
  summary(pairs(emm), infer = TRUE)
)
contrast_df
```

# Data Visualization

## Inactivation data facet by both target temp and strain

```{r}
df_dyn %>%
  ggplot(aes(x = time_min, y = log_cfu)) +
  geom_point() +
  facet_grid(strain~temp_target_c)
```

### Statistical comparison of 2 strains

```{r}
# Linear mixed model (paired/blocking by temp * time)
# Goal: Estimate difference while allowing each (temp,time) block to have its own baseline level (random intercept).
df_dyn_test <- df_dyn %>%
  mutate(cond  = interaction(temp_target_c, time_min, drop = TRUE))

# Model with strain-by-temperature fixed effect
fit <- lme(log_cfu ~ strain * temp_target_c, random = ~ 1 | cond,
           weights = varIdent(form = ~1 | temp_target_c),
           data = df_dyn_test)

anova(fit)

# Estimated strain contrasts within each target temperature
emm <- emmeans(fit, ~ strain | temp_target_c)
contrast_df <- as.data.frame(
  summary(pairs(emm), infer = TRUE)
)
contrast_df
plot_strain_temp <- ggplot(contrast_df, aes(x = estimate, y = temp_target_c)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = lower.CL, xmax = upper.CL), height = 0.15, linewidth = 0.9) +
  scale_x_continuous(breaks = seq(-2, 2, by = 0.1)) +
  geom_point(size = 3) +
  labs(x = expression("FDA_R2 - TX_2103 ("*log[10]*" CFU/g)"),
    y = expression("Target temperature ("*degree*C*")")) +
  theme_classic(base_size = 14) +
  theme(axis.text=element_text(size=14,color="black"))

plot_strain_temp 
```

```{r}
df_63 <- df_dyn %>%
  filter(temp_target_c == 63) %>%
  mutate(time_min = factor(time_min))

fit_63 <- lme(
  log_cfu ~ strain * time_min,
  random = ~ 1 | rep,
  weights = varIdent(form = ~1 | time_min),
  data = df_63
)

emm_63 <- emmeans(fit_63, ~ strain | time_min)

contrast_63 <- as.data.frame(
  summary(pairs(emm_63), infer = TRUE)
)

plot_strain_time <- ggplot(contrast_63, aes(x = as.numeric(as.character(time_min)), y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.8) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.3, linewidth = 0.9) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  labs(
    x = "Time (min)",
    y = expression("FDA_R2 - TX_2103 ("*log[10]*" CFU/g)")
  ) +
  theme_classic(base_size = 14)+
  theme(axis.text=element_text(size=14,color="black"))

combined_plot <- plot_strain_temp + plot_strain_time +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 16, face = "bold")
  )

combined_plot

ggsave(
  here("Plot", "Figure_strain_comparison_combined.tiff"),
  plot = combined_plot,
  width = 11,
  height = 4,
  units = "in",
  dpi = 600,
  compression = "lzw"
)
```

No significant different was observed between VP2 and VP3. Strain variable will be removed.

## **Inactivation data with strain combined**

```{r}
df_dyn %>%
  ggplot(aes(x = time_min, y = log_cfu)) +
  geom_point() +
  facet_grid(~temp_target_c)
```

## **time-temperature profile**

```{r}
df_temp <- df_comeup %>%
  group_by(temp_target_c, time_min) %>%
  summarise(temp_mean = mean(temp_recorded_c_ave),
            temp_sd = sd(temp_recorded_c_ave), .groups = "drop")

p_temp <- df_temp %>%
  ggplot(aes(
    x = time_min, y = temp_mean,
    color = temp_target_c,
    linetype = temp_target_c,
    group = temp_target_c
  )) +
  # Mean and SD ribbon
  geom_ribbon(aes(
    ymin = temp_mean - temp_sd,
    ymax = temp_mean + temp_sd,
    fill = temp_target_c
  ),
  alpha = 0.15, color = NA) +
  geom_line(size = 0.5) +
  geom_point(size = 1) +
  labs(
    y = bquote(Temperature~(degree*C)),
    x = "Time (min)",
    color = bquote(Target~Temperature~(degree*C)),
    linetype = bquote(Target~Temperature~(degree*C))
  ) +
  scale_color_manual(values = c("63" = "#ED0F54", "74" = "#52DD12", "98" = "#2998E2")) +
  scale_fill_manual(values  = c("63" = "#ED0F54", "74" = "#52DD12", "98" = "#2998E2")) +
  scale_linetype_manual(values = c("63" = "solid", "74" = "dashed", "98" = "longdash")) +
  scale_x_continuous(breaks = seq(0, 13, by = 1)) +
  scale_y_continuous(breaks = seq(20, 100, by = 10)) +
  coord_cartesian(xlim = c(0, 13), ylim = c(20, 100)) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.justification = "center",
    legend.margin = margin(t = 0, r = 0, b = 0, l = 0),
    legend.box.margin = margin(t = -8, r = 0, b = 0, l = 0),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 9),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)) +
  guides(fill = "none")

p_temp
ggsave(file = here("Plot", "ComeUp_Temp_Plot.tiff"), plot = p_temp, width = 6, height = 4, units = "in", dpi = 400)
```

# Data preprocessing for modeling

```{r}
# Left join the come up time df to get the measured temperature
df_dyn2 <- df_dyn %>%
  left_join(
    df_temp %>% dplyr::select(temp_target_c, time_min, temp_mean),
    by = c("temp_target_c", "time_min")
  ) %>%
  mutate(count = as.integer(round(cfu)))

saveRDS (df_dyn2, here("Data", "df_dyn2.RDS"))
```

## **Checking distribution of count**

```{r}
df_dyn2 <- readRDS (here("Data", "df_dyn2.RDS"))
# start with ggplot
ggplot(df_dyn2, aes(count)) +
  geom_histogram(binwidth = 10) +
  theme_light()

# Check if variance = mean
g <- df_dyn2 %>%
  group_by(temp_target_c, time_min, dilution, plated_ml) %>%
  summarise(
    n = n(),
    mean = mean(count, na.rm = TRUE),
    var = var(count,  na.rm = TRUE),
    vmr = var / mean, # variance-to-mean ratio
    .groups = "drop"
  )

summary(g$vmr)
quantile(g$vmr, c(.5, .75, .9), na.rm = TRUE)

# Variance estimation
g2 <- g %>%
  filter(mean > 0, vmr > 1) %>%
  mutate(size_hat = mean / (vmr - 1))

# confirm the distribution type with fitdistrplus package
fit_pois <- fitdist(df_dyn2$count, "pois")
fit_nb <- fitdist(df_dyn2$count, "nbinom")
plot(fit_pois)
gofstat(fit_pois)
plot(fit_nb)
gofstat(fit_nb)
exp(0.25)
# initial size estimation
summary(g2$size_hat)
quantile(g2$size_hat, c(0.25, 0.5, 0.75), na.rm = TRUE)
```

Variance-to-mean ratios far exceeding 1, violating Poisson assumptions, also, fitdist function confirms the selection of negative binomial. Therefore, a negative binomial likelihood was used for all subsequent model fitting.

# Functions for modeling

## **Model Fitting**

```{r}
get_nll <- function(p, this_data, known, times,
                    model_name = "Geeraerd") {
  
  # if (this_data$temp_target_c[1] == 74) {
  #   browser()
  # }
  
  all_pars <- as.list(c(p, known))
  
  # unlog transformation
  if ("logN0" %in% names(all_pars)) {
    all_pars$N0 <- 10^all_pars$logN0
    all_pars$logN0 <- NULL
  }
  if ("logNmin" %in% names(all_pars)) {
    all_pars$N_min <- 10^all_pars$logNmin
    all_pars$logNmin <- NULL
  }
  
  # Negative binomial distribution require size (dispersion) as parameter estimate
  size <- exp(all_pars$log_size)
  all_pars$log_size <- NULL
  
  # Temp profile for this target temperature condition
  this_temp <- this_data %>%
    dplyr::select(time, temperature) %>%
    distinct() %>%
    arrange(time)
  
  # Predicted data
  pred <- predict_inactivation(
    simulation_model = model_name,
    times = times,
    parms = all_pars,
    temp_profile = this_temp
  )$simulation %>%
    as.data.frame()
  
  # Observed data
  obs <- this_data %>%
    dplyr::select(time, count, dilution, plated_ml)
  
  merged <- merge(
    x = obs,
    y = pred[, c("time", "logN")],
    by = "time",
    all.x = TRUE,
    suffixes = c("_obs", "_mod")
  )
  
  # if (any(is.na(merged$logN)) || any(!is.finite(merged$logN))) return(Inf)
  
  # Predicted concentration from predicted log10 CFU/ml
  pred_N <- 10^merged$logN
  pred_count <- pred_N * merged$plated_ml * merged$dilution
  
  # if (any(!is.finite(pred_count)) || any(pred_count <= 0)) return(Inf)
  pred_count <- pmax(pred_count, 1e-12)
  
  # NB log-likelihood
  ll <- dnbinom(merged$count, mu = pred_count, size = size, log = TRUE)
  # if (any(!is.finite(ll))) {
  #   print("out")
  #   return(Inf)
  # } 
  
  # print(sum(ll))
  
  -sum(ll, na.rm = TRUE)
}

get_total_nll <- function(p, split_data, known, times_list,
                          model_name = "Geeraerd") {
  nlls <- mapply(
    FUN = function(dat, times) {
      get_nll(p = p, this_data = dat, known = known, times = times, 
              model_name = model_name)},
    dat = split_data,
    times = times_list,
    SIMPLIFY = TRUE
  )
  
  # if (any(!is.finite(nlls))) return(Inf)
  
  sum(nlls, na.rm = TRUE)
}

global_fit <- function(d, guess, known, model_name = "Geeraerd", upper = Inf, lower = -Inf, niter = 5000) {
  d2 <- d %>%
    mutate(
      time = as.numeric(time_min),
      temperature = as.numeric(temp_mean),
      time = ifelse(time < 2e-5, 2e-5, time),
      temp_target_c = as.factor(temp_target_c)
    )
  
  ## Split by target temperature
  split_data <- split(d2, d2$temp_target_c)
  
  ## Precompute times per condition
  times_list <- lapply(split_data, function(df) {
    df$time %>% unique() %>% sort() %>% na.omit()
  })
  
  modMCMC(f = get_total_nll, p = guess, split_data = split_data, 
          known = known, times_list = times_list, model_name = model_name, 
          upper = upper, lower = lower, niter = niter
  )
}
```

\##**Goodness-of-fit**

```{r}
compute_residuals_dyn <- function(data, model_name, bestpar, known_pars) {

  bp <- as.list(bestpar)
  size <- exp(as.numeric(bp$log_size))
  bp$log_size <- NULL
  
  # Unlog parameters if present
  if ("logN0" %in% names(bp)) {
    bp$N0 <- 10^bp$logN0
    bp$logN0 <- NULL
  }
  if ("logNmin" %in% names(bp)) {
    bp$N_min <- 10^bp$logNmin
    bp$logNmin <- NULL
  }
  
  
  d2 <- data %>%
    mutate(
      cond = as.factor(temp_target_c),
      time = round(as.numeric(time_min), 3),
      temperature = as.numeric(temp_mean)
    )
  
  preds <- d2 %>%
    split(.$cond) %>%
    map_df(function(df) {
      
      temp_prof <- df %>%
        dplyr::select(time, temperature) %>%
        distinct() %>%
        arrange(time)
      
      times_use <- sort(unique(df$time))
      
      sim <- quiet_predict_inactivation(
        simulation_model = model_name,
        times = times_use,
        parms = c(bp, as.list(known_pars)),
        temp_profile = temp_prof
      )$simulation %>%
        as.data.frame() %>%
        mutate(time = round(time, 3))
      
      data.frame(cond = df$cond, time = sim$time, m = sim$logN)
    }) %>%
    distinct(cond, time, .keep_all = TRUE)
  
  d2 %>%
    left_join(preds, by = c("cond", "time")) %>%
    mutate(pred_count = (10^m) * plated_ml * dilution,
           pred_count = ifelse(is.finite(pred_count) & pred_count > 0, pred_count, NA_real_))
}

# AIC and AICc calculation from log-likelihood
compute_fitstats <- function(res_df, bestpar, k, comm_df = NULL) {
  size <- exp(bestpar[["log_size"]])

  # If comm_df is provided, restrict to exactly those rows
  if (!is.null(comm_df)) {
    res_df <- res_df %>%
      dplyr::semi_join(
        comm_df %>% dplyr::distinct(cond, time, dilution, plated_ml, count),
        by = c("cond", "time", "dilution", "plated_ml", "count")
      )
  }

  ll <- dnbinom(
    x = res_df$count,
    mu = res_df$pred_count,
    size = size,
    log = TRUE
  )

  logLik <- sum(ll)
  n_all  <- nrow(res_df)

  AIC  <- 2 * k - 2 * logLik
  AICc <- AIC + (2 * k * (k + 1)) / (n_all - k - 1)

  list(logLik = logLik, AIC = AIC, AICc = AICc, n_used = n_all)
}
```

# Modeling

\##**Peleg Weibull** - Best fit

```{r}
df_dyn2 <- readRDS (here("Data", "df_dyn2.RDS"))

known_pe <- c()

guess_pe <- c(k_b = 0.1, n = 1.0, temp_crit = 50, logN0 = 6,log_size = log(1.9))

lower_pe <- c(k_b = 0.001, n = 0.1, temp_crit = 45, logN0 = 5, log_size = log(0.05))

upper_pe <- c(k_b = 20, n = 5, temp_crit = 65, logN0 = 7, log_size = log(20))

set.seed(5025)

invisible(capture.output(my_model_pe <- global_fit(
  d = df_dyn2, guess = guess_pe, known = known_pe, model_name = "Peleg",
  niter = 5000, lower = lower_pe, upper = upper_pe),type = "output"))

plot(my_model_pe)

summary(my_model_pe)

apply(my_model_pe$pars, 2, quantile, probs = c(0.025, 0.975))
```

### Decimal reduction time calculation

```{r}
temps <- c(63, 74, 90, 98)

pars <- as.data.frame(my_model_pe$pars)
pars$s <- exp(pars$log_size)
mean(pars$s)
sd(pars$s)
quantile(pars$s, probs = c(0.025, 0.975))
# compute b(T)
calc_b <- function(T, k_b, temp_crit){
  log1p(exp(k_b * (T - temp_crit)))
}

# compute D-value
calc_D <- function(b, n){
  (log(10) / b)^(1 / n)
}

# compute posterior samples
post_samples <- pars %>%
  mutate(
    b63 = calc_b(63, k_b, temp_crit),
    b74 = calc_b(74, k_b, temp_crit),
    b98 = calc_b(98, k_b, temp_crit),

    D63 = calc_D(b63, n),
    D74 = calc_D(b74, n),
    D98 = calc_D(b98, n)
  )

# summarize mean + 95% CrI
summary_table <- post_samples %>%
  summarise(
    b63_mean = mean(b63),
    b63_lwr = quantile(b63, 0.025),
    b63_upr = quantile(b63, 0.975),
    b63_sd = sd(b63),
    
    b74_mean = mean(b74),
    b74_lwr = quantile(b74, 0.025),
    b74_upr = quantile(b74, 0.975),
    b74_sd = sd(b74),
    
    b98_mean = mean(b98),
    b98_lwr = quantile(b98, 0.025),
    b98_upr = quantile(b98, 0.975),
    b98_sd = sd(b98),

    D63_mean = mean(D63),
    D63_lwr = quantile(D63, 0.025),
    D63_upr = quantile(D63, 0.975),
    D63_sd = sd(D63),

    D74_mean = mean(D74),
    D74_lwr = quantile(D74, 0.025),
    D74_upr = quantile(D74, 0.975),
    D74_sd = sd(D74),

    D98_mean = mean(D98),
    D98_lwr = quantile(D98, 0.025),
    D98_upr = quantile(D98, 0.975),
    D98_sd = sd(D98),
  )

summary_table
```

### Prediction of reduction

```{r}
### Predicted log10 reduction from model parameters only
#| message: false
#| warning: false

pars <- as.data.frame(my_model_pe$pars)

predict_reduction_one_draw <- function(profile_df, draw_row) {
  
  p <- as.list(draw_row)
  
  p$log_size <- NULL
  
  # Arbitrary starting value only used as a reference point.
  # It cancels out when calculating log10 reduction.
  p$N0 <- 10^6
  p$logN0 <- NULL
  
  temp_prof <- profile_df %>%
    dplyr::select(time = time_min, temperature = temp_mean) %>%
    dplyr::distinct() %>%
    dplyr::arrange(time)
  
  times_use <- seq(0, max(temp_prof$time), by = 0.01)
  
  sim <- suppressWarnings(
    quiet_predict_inactivation(
      simulation_model = "Peleg",
      times = times_use,
      parms = p,
      temp_profile = temp_prof
    )$simulation %>%
      as.data.frame()
  )
  
  sim %>%
    dplyr::mutate(
      log_reduction = logN[which.min(time)] - logN
    )
}

profiles_split <- split(df_temp, df_temp$temp_target_c)

reduction_all <- purrr::imap_dfr(profiles_split, function(prof, target_name) {
  
  purrr::map_dfr(seq_len(nrow(pars)), function(i) {
    
    predict_reduction_one_draw(
      profile_df = prof,
      draw_row = pars[i, ]
    ) %>%
      dplyr::mutate(
        draw = i,
        temp_target_c = target_name
      )
  })
})

reduction_table_1min <- reduction_all %>%
  dplyr::group_by(temp_target_c, time) %>%
  dplyr::summarise(
    mean_log_reduction = mean(log_reduction, na.rm = TRUE),
    lwr_log_reduction = quantile(log_reduction, 0.025, na.rm = TRUE),
    upr_log_reduction = quantile(log_reduction, 0.975, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::filter(abs(time - round(time)) < 1e-8) %>%
  dplyr::arrange(temp_target_c, time)

print(reduction_table_1min, n = Inf, width = Inf)
```

### Plot using the best parameter set with observed value overlaid

```{r}
df_dyn_fit <- df_dyn2 %>%
  mutate(
    temp_target_c = as.factor(temp_target_c),
    time = as.numeric(time_min),
    temperature = as.numeric(temp_mean)
  )

# un-log
fit_p_pe <- as.list(my_model_pe$bestpar)

if ("logN0" %in% names(fit_p_pe)) {
  fit_p_pe$N0 <- 10^fit_p_pe$logN0
  fit_p_pe$logN0 <- NULL
}

if ("logNmin" %in% names(fit_p_pe)) {
  fit_p_pe$N_min <- 10^fit_p_pe$logNmin
  fit_p_pe$logNmin <- NULL
}

if ("log_size" %in% names(fit_p_pe)) {
  fit_p_pe$log_size <- NULL
}

LOD <- 1.7

plots <- df_dyn_fit %>%
  split(.$temp_target_c) %>%
  purrr::map(~{
    temp_prof <- .x %>%
      dplyr::select(time, temperature) %>%
      distinct() %>%
      arrange(time)
    
    times_pred <- seq(0, max(.x$time, na.rm = TRUE), length.out = 200)
    
    pred <- quiet_predict_inactivation(
      simulation_model = "Peleg",
      times = times_pred,
      parms = c(fit_p_pe, as.list(known_pe)),
      temp_profile = temp_prof
    )

    sim <- pred$simulation %>%
      as.data.frame() %>%
      mutate(logN_clip = pmax(logN, LOD))

    ggplot() +
      geom_hline(yintercept = LOD, linetype = "dotted", color = "gray50") +
      geom_line(data = sim, aes(x = time, y = logN), linewidth = 0.9) +
      geom_point(data = .x, aes(x = time, y = log_cfu), size = 1.6) +
      labs(
        title = paste0("Target ", as.character(unique(.x$temp_target_c)), "°C"),
        x = "time (min)",
        y = expression(Log[10]~"CFU/mL")
      ) +
      coord_cartesian(ylim = c(0, 6.5)) +
      theme_classic()
  })

cowplot::plot_grid(plotlist = plots)
```

### Goodness-of-fit check

```{r}
df_resid_pe <- compute_residuals_dyn(
  data = df_dyn2,
  model_name = "Peleg",
  bestpar = my_model_pe$bestpar,
  known_pars = known_pe
)


# fit stats
fitstats_pe <- compute_fitstats(
  res_df = df_resid_pe,
  bestpar = my_model_pe$bestpar,
  k = 4
)

fitstats_pe$AIC
fitstats_pe$AICc
```

### Bacterial reduction estimation given different initial conc.

```{r eval=FALSE, echo=FALSE}
## Set new initial concentration
logN0_new <- 3
N0_new <- 10^logN0_new
pars <- as.data.frame(my_model_pe$pars)
# prediction function: use df_temp directly
predict_one_draw_one_profile <- function(profile_df, draw_row, model_name = "Peleg") {
  
  # parameter list from one posterior draw
  p <- as.list(draw_row)
  
  # remove observation-model parameter
  p$log_size <- NULL
  
  # replace fitted initial conc. with 3 log
  p$N0 <- N0_new
  p$logN0 <- NULL
  
  # prepare temp profile in the format required by predict_inactivation()
  temp_prof <- profile_df %>%
    dplyr::select(time = time_min, temperature = temp_mean) %>%
    distinct() %>%
    arrange(time)
  
  # prediction times wanted in output
  times_use <- seq(0, 8, by = 0.01)
  
  sim <- suppressWarnings(
    quiet_predict_inactivation(
    simulation_model = model_name,
    times = times_use,
    parms = p,
    temp_profile = temp_prof
  )$simulation %>%
    as.data.frame()
)
  sim
}

# split by target temperature
profiles_split <- df_temp %>%
  split(.$temp_target_c)

# predict for all posterior draws and all target profiles

pred_all <- imap_dfr(profiles_split, function(prof, target_name) {
  
  map_dfr(seq_len(nrow(pars)), function(i) {
    
    sim <- predict_one_draw_one_profile(
      profile_df = prof,
      draw_row = pars[i, ],
      model_name = "Peleg"
    )
    
    sim %>%
      mutate(
        draw = i,
        temp_target_c = target_name
      )
  })
})

# summarize posterior predictions
pred_sum <- pred_all %>%
  group_by(temp_target_c, time) %>%
  summarise(
    mean_logN = mean(logN, na.rm = TRUE),
    sd_logN = sd(logN, na.rm = TRUE),
    lwr_95 = quantile(logN, 0.025, na.rm = TRUE),
    upr_95 = quantile(logN, 0.975, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    temp_target_c = factor(as.character(temp_target_c), levels = c("63", "74", "98")),
    mean_log_reduction = logN0_new - mean_logN,
    lwr_log_reduction = logN0_new - upr_95,
    upr_log_reduction = logN0_new - lwr_95
  ) %>%
  arrange(temp_target_c, time) %>%
  mutate(
    mean_logN = pmax(mean_logN, 0),
    lwr_95 = pmax(lwr_95, 0),
    upr_95 = pmax(upr_95, 0)
  )

saveRDS(pred_sum, here("Result", "prediction_summary.RDS"))

p_pred <- ggplot(pred_sum,
                 aes(x = time,
                     y = mean_logN,
                     color = factor(temp_target_c),
                     fill = factor(temp_target_c))) +
  geom_ribbon(aes(ymin = lwr_95, ymax = upr_95), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.1) +
  geom_point(size = 0.5) +
  scale_color_manual(values = c("63" = "#ED0F54", "74" = "#52DD12", "98" = "#2998E2")) +
  scale_fill_manual(values = c("63" = "#ED0F54", "74" = "#52DD12", "98" = "#2998E2")) +
  scale_x_continuous(breaks = seq(0, 8, by = 1)) +
  scale_y_continuous(limits = c(0, logN0_new), expand = c(0,0)) +
  labs(
    x = "Time (min)",
    y = expression("Predicted " * log[10] * " CFU/g Concentration"),
    color = "Target temperature (°C)",
    fill = "Target temperature (°C)",
  ) +
  theme_bw()

p_pred
ggsave(file = here("Plot", "bacterial_reduction_prediction_Plot.tiff"), plot = p_pred, width = 6, height = 4, units = "in", dpi = 400)

# every min from the smooth simulation
pred_table_min <- pred_sum %>%
  filter(abs(time - round(time / 1) * 1) < 1e-8) %>%
  mutate(
    mean_log_reduction = logN0_new - mean_logN,
    lwr_log_reduction = logN0_new - upr_95,
    upr_log_reduction = logN0_new - lwr_95
  ) %>%
  arrange(temp_target_c, time)

pred_table_min

saveRDS(pred_table_2min, here("Result", "logN_prediction_table_2min_peleg.RDS"))
```

## Plot for publication

```{r}
make_interv_pe <- function(m, data, known, model = "Peleg") {
  
  p <- unlist(m)
  p <- p[!names(p) %in% c("log_size", "i")]
  p["N0"] <- 10^p["logN0"]
  
  data %>%
    split(.$temp_target_c) %>%
    map(~{
      dat <- .x %>%
        mutate(
          time = as.numeric(time_min),
          temperature = as.numeric(temp_mean)
        )
      
      temp_prof <- dat %>%
        dplyr::select(time, temperature) %>%
        distinct() %>%
        arrange(time)
      
      times_pred <- seq(0, max(dat$time, na.rm = TRUE), length.out = 200)
      
      predict_inactivation(
        simulation_model = model,
        times = times_pred,
        parms = c(p, known),
        temp_profile = temp_prof
      )$simulation %>%
        as.data.frame()
    }) %>%
    imap_dfr(~ mutate(.x, temp_target_c = .y))
}

## Posterior predictions
MCMC_preds_pe <- my_model_pe$pars %>%
  as_tibble() %>%
  slice_sample(n = 1000, replace = TRUE) %>%
  mutate(i = row_number()) %>%
  split(.$i) %>%
  purrr::map_dfr(
    ~ {
      invisible(capture.output(
        pred <- make_interv_pe(.x, data = df_dyn2, known = known_pe, model = "Peleg"),
        type = "output"
      ))
      pred
    },
    .id = "i"
  ) %>%
  summarise(
    m = median(logN, na.rm = TRUE),
    q2.5 = quantile(logN, 0.025, na.rm = TRUE),
    q97.5 = quantile(logN, 0.975, na.rm = TRUE),
    .by = c("temp_target_c", "time")
  )
```

```{r}
## rescale temperature to match CFU axis
temp_scaled <- df_temp %>%
  mutate(temp_target_c=factor(temp_target_c),
         time=as.numeric(time_min),
         scal_temp=(temp_mean-20)/(100-20)*6)

LOD <- 1.7  # limit of detection line

make_panel <- function(Ttarget, xlim=NULL, xbreaks=NULL, 
                       show_left_axis=FALSE,
                       show_right_axis=FALSE,
                       show_temp_title=FALSE){
  
  ## model predictions (clip at 0 for plotting)
  pred_sub <- MCMC_preds_pe %>%
    filter(temp_target_c==Ttarget) %>%
    mutate(m_plot=pmax(m,0),
           q2.5_plot=pmax(q2.5,0),
           q97.5_plot=pmax(q97.5,0))
  
  ## observed data
  dat_sub <- df_dyn2 %>%
    filter(temp_target_c==Ttarget) %>%
    mutate(time=as.numeric(time_min),
           logN=as.numeric(log_cfu),
           strain=factor(strain))
  
  ## temperature profile (already rescaled)
  temp_sub <- temp_scaled %>%
    filter(temp_target_c==Ttarget)
  
  ggplot() +
    ## uncertainty ribbon (95% CI)
    geom_ribbon(data=pred_sub,
                aes(x=time,ymin=q2.5_plot,ymax=q97.5_plot),
                fill="gray80",alpha=0.8) +
    
    ## median prediction line
    geom_line(data=pred_sub,
              aes(x=time,y=m_plot),
              linewidth=0.9,color="black") +
    
    ## observed points (two strains as shapes)
    geom_point(data=dat_sub,
               aes(x=time,y=logN,shape=strain),
               size=1.8,stroke=0.5,color="black",fill="black") +
    
    ## temperature curve (scaled to same axis)
    geom_line(data=temp_sub,
              aes(x=time,y=scal_temp),
              linetype="longdash",linewidth=0.6,color="gray35") +
    
    ## LOD horizontal line
    geom_hline(yintercept=LOD,
               linetype="dotted",linewidth=0.6,color="gray45") +
    
    ## shape mapping for strains
    scale_shape_manual(values=c(16,2)) +
    
    ## primary + optional secondary axis
    scale_y_continuous(
      name=if(show_left_axis) expression(Log[10]~"CFU/g") else NULL, # left axis only for first panel
      breaks=seq(0,6,1),
      sec.axis=if(show_right_axis) sec_axis(                         # right axis only for last panel
        ~./6*80+20,
        breaks=seq(20,100,20),
        name=if(show_temp_title) "Temperature (°C)" else NULL
      ) else waiver()
    ) +
    
    ## x-axis
    scale_x_continuous(breaks=xbreaks) +
    
    ## limits
    coord_cartesian(xlim=xlim,ylim=c(0,6)) +
    
    labs(x= "Time (min)",title=paste0("Target ",as.character(Ttarget),"°C")) +
    
    ## styling
    theme_bw(base_size=12) +
    theme(
      plot.title=element_text(size=10,face="bold",hjust=0.5),
      axis.text.x=element_text(size=12,color="black"),
      ## left axis visibility control
      axis.text.y.left=if(show_left_axis) element_text(size=12,color="black") else element_blank(),
      axis.ticks.y.left=if(show_left_axis) element_line(color="black") else element_blank(),
      axis.title.y.left=if(show_left_axis) element_text(size=12) else element_blank(),
      
      ## right axis visibility control
      axis.text.y.right=if(show_right_axis) element_text(size=12,color="gray25") else element_blank(),
      axis.ticks.y.right=if(show_right_axis) element_line(color="gray25") else element_blank(),
      axis.title.y.right=if(show_right_axis && show_temp_title)
        element_text(size=12,margin=margin(l=6)) else element_blank(),
      
      ## remove grids + legend
      panel.grid.minor=element_blank(),
      panel.grid.major=element_blank(),
      panel.border=element_rect(linewidth=0.6,color="black"),
      legend.position="none"
    )
}

## build panels
p1 <- make_panel("63", xlim=c(0,13), xbreaks=seq(0,12,2), show_left_axis=TRUE)
p2 <- make_panel("74", xlim=c(0,13), xbreaks=seq(0,12,2))
p3 <- make_panel("98", xlim=c(0,13), xbreaks=seq(0,12,2), show_right_axis=TRUE, show_temp_title=TRUE)

## combine (tight layout)
p_all <- p1 + p2 + p3 + plot_layout(ncol=3)

p_all

## save
ggsave(filename=here("Plot","Vibrio_Oys_PE_Plot.tiff"),
       plot=p_all, width=10, height=4, units="in",
       dpi=1000,bg="white")
```

## **Geeraerd**

```{r}
model_name <- "Geeraerd"
known_ge_1 <- c(temp_ref = 63)
lower_ge_1 <- c(D_R = .3, z = 2, logN0 = 5, logNmin = -1, log_size = log(0.05), C_c0 = 1e-6)
upper_ge_1 <- c(D_R = 6, z = 15, logN0 = 7, logNmin = 1.7, log_size = log(20), C_c0 = 1e3)

# Initial guess from the inactivation data plot

guess_ge_1 <- c(D_R = 2.4, z = 6, logN0 = 6, logNmin = 0, log_size = log(1.9), C_c0 = 1)

set.seed(5025)
my_model_ge_1 <- global_fit(df_dyn2, guess_ge_1, known_ge_1, niter = 5000,
                            lower = lower_ge_1, 
                            upper = upper_ge_1)

saveRDS(my_model_ge_1, here("Result", "model_Geeraerd_1.RDS"))

plot(my_model_ge_1)
apply(my_model_ge_1$pars, 2, quantile, probs = c(0.025, 0.975))
summary(my_model_ge_1)
```

From the summary, the logNmin is unidentifiable in the censored-likelihood formulation, so MCMC pins it to the boundary 0. Let's try put N_min as the known value 10\^0

```{r}
model_name <- "Geeraerd"
known_ge <- c(temp_ref = 63, N_min = 10^0)
lower_ge <- c(D_R = .3, z = 2, logN0 = 5, log_size = log(0.05), C_c0 = 1e-6)
upper_ge <- c(D_R = 6, z = 15, logN0 = 7, log_size = log(20), C_c0 = 1e3)
# Initial guess from the inactivation data plot
guess_ge <- c(D_R = 2.4, z = 6, logN0 = 6, log_size = log(1.9), C_c0 = 1)

my_model_ge <- global_fit(df_dyn2, guess_ge, known_ge, niter = 5000,
                       lower = lower_ge,
                       upper = upper_ge)

saveRDS(my_model_ge, here("Result", "model_Geeraerd.RDS"))

plot(my_model_ge)

summary(my_model_ge)
```

## **Geeraerd with no shoulder**

```{r}
model_name <- "Geeraerd"
known_ge_ns <- c(temp_ref = 63, N_min = 10^0, C_c0 = 1e-3)
lower_ge_ns <- c(D_R = .3, z = 2, logN0 = 5, log_size = log(0.05))
upper_ge_ns <- c(D_R = 6, z = 15, logN0 = 7, log_size = log(20))
# Initial guess from the inactivation data plot
guess_ge_ns <- c(D_R = 2.4, z = 6, logN0 = 6, log_size = log(1.9))


set.seed(5025)
my_model_ge_ns <- global_fit(df_dyn2, guess_ge_ns, known_ge_ns, niter = 5000,
                       lower = lower_ge_ns,
                       upper = upper_ge_ns)

saveRDS(my_model_ge_ns, here("Result", "model_Geeraerd_no_shoulder.RDS"))

plot(my_model_ge_ns)

summary(my_model_ge_ns)
```

### Plot using the best parameter set with observed value overlaid (with shoulder)

```{r}
# un-log
fit_p_ge <- as.list(my_model_ge$bestpar)

if ("logN0" %in% names(fit_p_ge)) {
  fit_p_ge$N0 <- 10^fit_p_ge$logN0
  fit_p_ge$logN0 <- NULL
}

if ("logNmin" %in% names(fit_p_ge)) {
  fit_p_ge$N_min <- 10^fit_p_ge$logNmin
  fit_p_ge$logNmin <- NULL
}

if ("log_size" %in% names(fit_p_ge)) {
  fit_p_ge$log_size <- NULL
}

plots <- df_dyn_fit %>%
  split(.$temp_target_c) %>%
  purrr::map(~{
    temp_prof <- .x %>%
      dplyr::select(time, temperature) %>%
      distinct() %>%
      arrange(time)
    
times_pred <- seq(0, max(.x$time, na.rm = TRUE), length.out = 200)

pred <- quiet_predict_inactivation(
  simulation_model = "Geeraerd",
  times = times_pred,
  parms = c(fit_p_ge, as.list(known_ge)),
  temp_profile = temp_prof
  )

  sim <- pred$simulation %>%
    as.data.frame() %>%
    mutate(logN_clip = pmax(logN, LOD))

  ggplot() +
    geom_hline(yintercept = LOD, linetype = "dotted", color = "gray50") +
    geom_line(data = sim, aes(x = time, y = logN), linewidth = 0.9) +
    geom_point(data = .x, aes(x = time, y = log_cfu), size = 1.6) +
    labs(
      title = paste0("Target ", as.character(unique(.x$temp_target_c)), "°C"),
      x = "time (min)",
      y = expression(Log[10]~"CFU/mL")
    ) +
    coord_cartesian(ylim = c(0, 6.5)) +
    theme_classic()
  })

cowplot::plot_grid(plotlist = plots)
```

### Plot using the best parameter set with observed value overlaid (with no shoulder)

```{r}
# un-log
fit_p_ge_ns <- as.list(my_model_ge_ns$bestpar)

if ("logN0" %in% names(fit_p_ge_ns)) {
  fit_p_ge_ns$N0 <- 10^fit_p_ge_ns$logN0
  fit_p_ge_ns$logN0 <- NULL
}

if ("logNmin" %in% names(fit_p_ge_ns)) {
  fit_p_ge_ns$N_min <- 10^fit_p_ge_ns$logNmin
  fit_p_ge_ns$logNmin <- NULL
}

if ("log_size" %in% names(fit_p_ge_ns)) {
  fit_p_ge_ns$log_size <- NULL
}

plots <- df_dyn_fit %>%
  split(.$temp_target_c) %>%
  purrr::map(~{
    temp_prof <- .x %>%
      dplyr::select(time, temperature) %>%
      distinct() %>%
      arrange(time)
    
times_pred <- seq(0, max(.x$time, na.rm = TRUE), length.out = 200)

pred <- quiet_predict_inactivation(
  simulation_model = "Geeraerd",
  times = times_pred,
  parms = c(fit_p_ge_ns, as.list(known_ge_ns)),
  temp_profile = temp_prof
  )

  sim <- pred$simulation %>%
    as.data.frame() %>%
    mutate(logN_clip = pmax(logN, LOD))

  ggplot() +
    geom_hline(yintercept = LOD, linetype = "dotted", color = "gray50") +
    geom_line(data = sim, aes(x = time, y = logN), linewidth = 0.9) +
    geom_point(data = .x, aes(x = time, y = log_cfu), size = 1.6) +
    labs(
      title = paste0("Target ", as.character(unique(.x$temp_target_c)), "°C"),
      x = "time (min)",
      y = expression(Log[10]~"CFU/mL")
    ) +
    coord_cartesian(ylim = c(0, 6.5)) +
    theme_classic()
  })

cowplot::plot_grid(plotlist = plots)
```

### Goodness-of-fit check (with shoulder)

```{r}
df_resid_ge <- compute_residuals_dyn(
  data = df_dyn2,
  model_name = "Geeraerd",
  bestpar = my_model_ge$bestpar,
  known_pars = known_ge
)


# fit stats
fitstats_ge <- compute_fitstats(
  res_df = df_resid_ge,
  bestpar = my_model_ge$bestpar,
  k = 4
)

fitstats_ge$AIC
fitstats_ge$AICc
```

### Goodness-of-fit check (with no shoulder)

```{r}
df_resid_ge_ns <- compute_residuals_dyn(
  data = df_dyn2,
  model_name = "Geeraerd",
  bestpar = my_model_ge_ns$bestpar,
  known_pars = known_ge_ns
)


# fit stats
fitstats_ge_ns <- compute_fitstats(
  res_df = df_resid_ge_ns,
  bestpar = my_model_ge_ns$bestpar,
  k = 4
)

fitstats_ge_ns$AIC
fitstats_ge_ns$AICc
```

## **Log-linear (Bigelow)**

```{r, warning=FALSE}
#| warning: false

known_b <- c(temp_ref = 63)

guess_b <- c(D_R = 2.5, z = 6, logN0 = 6, log_size = log(1.9))

set.seed(5025)
my_model_b <- global_fit(df_dyn2, guess_b, known_b, model_name ="Bigelow",
                       niter = 5000,
                       lower = c(D_R = 2, z = 5, logN0 = 5, log_size = log(0.05)),
                       upper = c(D_R = 20, z = 15, logN0 = 7, log_size = log(20)))

plot(my_model_b)  # check convergence

summary(my_model_b)
```

### Plot using the best parameter set with observed value overlaid

```{r, eval=FALSE, warning=FALSE}
# un-log
fit_p_b <- as.list(my_model_b$bestpar)

if ("logN0" %in% names(fit_p_b)) {
  fit_p_b$N0 <- 10^fit_p_b$logN0
  fit_p_b$logN0 <- NULL
}

if ("logNmin" %in% names(fit_p_b)) {
  fit_p_b$N_min <- 10^fit_p_b$logNmin
  fit_p_b$logNmin <- NULL
}

if ("log_size" %in% names(fit_p_b)) {
  fit_p_b$log_size <- NULL
}

plots <- df_dyn_fit %>%
  split(.$temp_target_c) %>%
  purrr::map(~{
    temp_prof <- .x %>%
      dplyr::select(time, temperature) %>%
      distinct() %>%
      arrange(time)
    
times_pred <- seq(0, max(.x$time, na.rm = TRUE), length.out = 200)

pred <- quiet_predict_inactivation(
  simulation_model = "Bigelow",
  times = times_pred,
  parms = c(fit_p_b, as.list(known_b)),
  temp_profile = temp_prof
  )

  sim <- pred$simulation %>%
    as.data.frame() %>%
    mutate(logN_clip = pmax(logN, LOD))

  ggplot() +
    geom_hline(yintercept = LOD, linetype = "dotted", color = "gray50") +
    geom_line(data = sim, aes(x = time, y = logN), linewidth = 0.9) +
    geom_point(data = .x, aes(x = time, y = log_cfu), size = 1.6) +
    labs(
      title = paste0("Target ", as.character(unique(.x$temp_target_c)), "°C"),
      x = "time (min)",
      y = expression(Log[10]~"CFU/mL")
    ) +
    coord_cartesian(ylim = c(0, 6.5)) +
    theme_classic()
  })

cowplot::plot_grid(plotlist = plots)
```

### Goodness-of-fit check

```{r, eval=FALSE, warning=FALSE}
df_resid_b <- compute_residuals_dyn(
  data = df_dyn2,
  model_name = "Bigelow",
  bestpar = my_model_b$bestpar,
  known_pars = known_b
)

# NA prediction was produced
df_resid_b %>%
  dplyr::filter(!is.finite(pred_count) | is.na(pred_count) | pred_count <= 0)
```
